Visual Odometry for Autonomous Underwater Vehicles

### 2013-03-21 paper.tex
CHANGE: results updated

Visual odometers for AUVs tend to use this pipeline \cite{Botelho2009} and adapt every stage options (such us keypoint detector type and feature matching method) to they needs. Common feature detectors used for real-time visual odometry include Harris corners \cite{Harris1988} \cite{Nister2006}, FAST \cite{Rosten2006} \cite{Huang2011} or SIFT \cite{Lowe2004} \cite{Botelho2009} features, which have a low computation cost and are robust against small viewpoint changes. Wide range of approaches are involved in the process of matching feature across frames, \cite{Geiger2011} propose a circle match between stereo image pairs of two consecutive frames and uses non-maxima-suppression techniques \cite{Geiger2010} to reduce the number of correspondences. RANSAC-based methods \cite{Nister2004} \cite{Johnson2008} and graph-based consistency algorithms \cite{Howard2008} have also been proven to robustly match features across frames.

Previous to the motion estimation stage, keypoint bucketing technique \cite{Zhang1995} has been shown to help reduce the inlier reprojection errors. Finally, motion estimation process can be solved through different methods. Using the closed-form solution to least-squares problem of absolute orientation \cite{Horn1987}, it is possible to directly minimize the Euclidean distance between matched features in order to compute the rigid-body transformation of the camera frame \cite{Huang2011}. Several ground-based visual odometers do not use 3D distances but implement the method of minimizing the pixel reprojection error \cite{Howard2008} \cite{Geiger2011}.

Other AUVs localization algorithms are based on mosaics or structured enviroments and therefore assumes prior knowledge of the area to explore \cite{Garcia2001} \cite{Gracias2003} \cite{Carreras2003}.

@@

fovis uses the FAST feature detector \cite{Rosten2005FPA,Rosten2006MLF} with an adaptive threshold on three gaussian pyramid levels. The number of features is reduced taking the best features of each cell of a regular grid. The feature descriptor is an intensity normalized 9x9 pixel window centered on the feature location. The bottom right pixel is left out to get better memory alignment for rapid similarity measure computation using SIMD instructions.

@@

A comparison of linear velocity estimates to ground truth can be found in Figure \ref{error-plots-viso2} for libviso2 and \ref{error-plots-fovis} for fovis. Upper graphics refer to experiments in laboratory while the ones of the bottom refer to harbour tests. Linear velocity is shown in left graphics for both odometry estimation (blue) and ground truth (green), together with its error drawn in red. Finally, on right plot, a comparison of the angular velocity for odometry estimation (blue) and grund truth is also presented. The error in angular velocity is insignificant thus not shown in the figures.

This results are numerically summarized in Table \ref{results-cirs} for laboratory experiments and Table \ref{results-roses} for harbour experiments.

@@

\begin{figure*}
  \begin{center}
    \subfloat[]{
      \begin{tabular}{c}
        \includegraphics[width=0.45\linewidth]{aux/fovis-cirs-timefixed-run3-vt.pdf}\\
        \includegraphics[width=0.45\linewidth]{aux/fovis-roses-vt.pdf}
      \end{tabular}
    }
    \subfloat[]{
      \begin{tabular}{c}
        \includegraphics[width=0.45\linewidth]{aux/fovis-cirs-timefixed-run3-vyaw.pdf}\\
        \includegraphics[width=0.45\linewidth]{aux/fovis-roses-vyaw.pdf}
      \end{tabular}
    }
  \end{center}
  \caption{Linear (a) and angular (b) velocity estimates compared to ground truth for the laboratory (top) and the harbour (bottom) experiments for fovis.}
  \label{error-plots-fovis}
\end{figure*}

TODO Brief description of parameter tuning

\begin{table}[!t]
  \renewcommand{\arraystretch}{1.3}
  \caption{Comparison of translational and rotational errors for the laboratory experiments.}
  \label{results-cirs}
  \centering
  \begin{tabular}{|l|l|l|l|} \hline
    \textbf{Algorithm} & \textbf{Translation} & \textbf{Rotation} & \textbf{Runtime}\\ \hline
    libviso2 & 0.978\% & 0.001739 [deg/m] & TODO \\ \hline
    fovis    & 0.601\% & 0.006311 [deg/m] & TODO \\ \hline
  \end{tabular} 
\end{table}

\begin{table}[!t]
  \renewcommand{\arraystretch}{1.3}
  \caption{Comparison of translational and rotational errors for the harbour experiments.}
  \label{results-roses}
  \centering
  \begin{tabular}{|l|l|l|l|} \hline
    \textbf{Algorithm} & \textbf{Translation} & \textbf{Rotation} & \textbf{Runtime}\\ \hline
    libviso2 & 0.833\% & 0.007913 [deg/m] & TODO \\ \hline
    fovis    & 1.049\% & 0.003122 [deg/m] & TODO \\ \hline
  \end{tabular} 
\end{table}

TODO Runtime comparison (table)

TODO Keypoints detection comparison (table)

TODO describe difficulties / explain obvious errors in the data etc.

>>>>

Visual odometers for AUVs tend to use this pipeline \cite{Botelho2009} and adapt every stage options (such us keypoint detector type and feature matching method) to they needs. Common feature detectors used for real-time visual odometry include Harris corners \cite{Harris1988,Nister2006}, FAST \cite{Rosten2006,Huang2011} or SIFT \cite{Lowe2004,Botelho2009} features, which have a low computation cost and are robust against small viewpoint changes. Wide range of approaches are involved in the process of matching feature across frames, \cite{Geiger2011} propose a circle match between stereo image pairs of two consecutive frames and uses non-maxima-suppression techniques \cite{Geiger2010} to reduce the number of correspondences. RANSAC-based methods \cite{Nister2004,Johnson2008} and graph-based consistency algorithms \cite{Howard2008} have also been proven to robustly match features across frames.

Previous to the motion estimation stage, keypoint bucketing technique \cite{Zhang1995} has been shown to help reduce the inlier reprojection errors. Finally, motion estimation process can be solved through different methods. Using the closed-form solution to least-squares problem of absolute orientation \cite{Horn1987}, it is possible to directly minimize the Euclidean distance between matched features in order to compute the rigid-body transformation of the camera frame \cite{Huang2011}. Several ground-based visual odometers do not use 3D distances but implement the method of minimizing the pixel reprojection error \cite{Howard2008,Geiger2011}.

Other AUVs localization algorithms are based on mosaics or structured enviroments and therefore assumes prior knowledge of the area to explore \cite{Garcia2001,Gracias2003,Carreras2003}.

@@

fovis uses the FAST feature detector \cite{Rosten2005FPA,Rosten2006} with an adaptive threshold on three gaussian pyramid levels. The number of features is reduced taking the best features of each cell of a regular grid. The feature descriptor is an intensity normalized 9x9 pixel window centered on the feature location. The bottom right pixel is left out to get better memory alignment for rapid similarity measure computation using SIMD instructions.

@@

A comparison of linear velocity estimates to ground truth can be found in Figure \ref{error-plots-viso2} for libviso2 and \ref{error-plots-fovis} for fovis. Upper graphics refer to experiments in laboratory while the ones of the bottom refer to harbour tests. Linear velocity is shown in left graphics for both odometry estimation (blue) and ground truth (green), together with its error drawn in red. Finally, on right plot, a comparison of angular velocity for odometry estimation (blue) and grund truth (green) is also presented. The error in angular velocity has a zero mean thus not shown in the figures. No differences can be highlighted comparing angular velocity error for libviso2 and fovis in both the laboratory and harbour conditions. However, when linear velocities are examined, fovis presents better performance when executed in laboratory enviroment.

These results are numerically summarized in Table \ref{results-errors} for laboratory and harbour experiments. Translation column indicates the mean average translation error that suffers the odometer in percent, which is the number of meters diverted for every 100 meters of trajectory. Furthermore, the rotation column indicates how many degrees the odometer is deviated per meter, relative to ground truth.

The results presented in Table \ref{results-errors} demostrates that libviso2 has similar mean translation error for both enviroments, but differs significantly in rotation error. By refering to Table \ref{results-runtime}, the mean number of inliers per frame is larger in the laboratory than in the harbour (where environmental conditions are worse). It means libviso2 presents a strong performance in the linear motion estimation when the number of inliers is reduced.

Results for fovis odometer are substantially different in translation error comparing laboratory and harbour conditions. By examining the number of inliers of the two odometers shown in Table \ref{results-runtime} it is easy to see that fovis is significantly less robust in harbour enviroments where the quality of the images is lower (see Figure \ref{sample-images}). Furthermore, unlike libviso2, fovis presents a better performance when estimating the angular motion in harbour conditions. This improvement over libviso2 is due to the method of keyframe selection based on the number of matches introduced by fovis. This technique has a positive impact when the vehicle experiences pure rotational movements, as it can keep the reference frame for longer and therefore reduce the drift significally.

Runtime column shown in Table \ref{results-runtime} indicates the mean algorithm execution time per frame. For both libviso2 and fovis odometers the runtime is clearly related to the number of inliers to be processed by the motion estimation algorithm, thus large number of inliers increases processing time.

@@

\begin{figure*}[!h]
  \begin{center}
    \subfloat[]{
      \begin{tabular}{c}
        \includegraphics[width=0.45\linewidth]{aux/fovis-cirs-timefixed-run3-vt.pdf}\\
        \includegraphics[width=0.45\linewidth]{aux/fovis-roses-vt.pdf}
      \end{tabular}
    }
    \subfloat[]{
      \begin{tabular}{c}
        \includegraphics[width=0.45\linewidth]{aux/fovis-cirs-timefixed-run3-vyaw.pdf}\\
        \includegraphics[width=0.45\linewidth]{aux/fovis-roses-vyaw.pdf}
      \end{tabular}
    }
  \end{center}
  \caption{Linear (a) and angular (b) velocity estimates compared to ground truth for the laboratory (top) and the harbour (bottom) experiments for fovis.}
  \label{error-plots-fovis}
\end{figure*}

\begin{table}[!t]
  \renewcommand{\arraystretch}{1.3}
  \caption{Comparison of translational and rotational errors for laboratory and harbour experiments.}
  \label{results-errors}
  \centering
  \begin{tabular}{|l|l|c|c|} \hline
    \textbf{Enviroment} & \textbf{Algorithm} & \textbf{Translation} & \textbf{Rotation}\\ \hline
    \multirow{2}{*}{Laboratory} & libviso2 & 0.978\% & 0.001739 [deg/m] \\
                                & fovis    & 0.601\% & 0.006311 [deg/m] \\ \hline
    \multirow{2}{*}{Harbour}    & libviso2 & 0.833\% & 0.007913 [deg/m] \\
                                & fovis    & 1.049\% & 0.003122 [deg/m] \\ \hline
  \end{tabular} 
\end{table}

\begin{table}[!t]
  \renewcommand{\arraystretch}{1.3}
  \caption{Comparison of number of inliear and runtime for laboratory and harbour experiments.}
  \label{results-runtime}
  \centering
  \begin{tabular}{|l|l|c|c|} \hline
    \textbf{Enviroment} & \textbf{Algorithm} & \textbf{Num. inliers} & \textbf{Runtime}\\ \hline
    \multirow{2}{*}{Laboratory} & libviso2 & 366 & 0.106 [s] \\
                                & fovis    & 500 & 0.246 [s] \\ \hline
    \multirow{2}{*}{Harbour}    & libviso2 & 137 & 0.086 [s] \\
                                & fovis    & 64 & 0.058 [s] \\ \hline
  \end{tabular} 
\end{table}
